\documentclass[a4paper]{report}

% Import packages
\usepackage[latin1]{inputenc}
\usepackage{color}
\usepackage{psboxit}
\usepackage{verbatim}
\usepackage{itools}

% Document header
\title{\tt itools}
\author{Juan David Ibáñez Palomar\thanks{Permission is granted to copy,
distribute and/or modify this document under the terms of the GNU Free
Documentation License, Version 1.2 or any later version published by the
Free Software Foundation; with no Invariant Sections, no Front-Cover Texts,
and no Back-Cover Texts.  There is a copy of the license at
http://www.gnu.org/copyleft/fdl.html}\\
\small \texttt{jdavid@itaapy.com}}


\begin{document}

\maketitle
\tableofcontents


\input{intro.tex}
\input{uri.tex}
\input{resources.tex}
\input{handlers.tex}
\input{stl.tex}



\chapter{Conclusion}

This article has presented three core features provided by {\tt itools}:
{\tt itools.uri}, {\tt itools.resources} and {\tt itools.handlers}. Many
things have been left aside to keep the article short: the description
of the possibilities of the XML handler, or a deeper discussion about how
to build your own handlers.

We hope you have found these tools interesting enough to get a closer look.

\section{In the real world}

In spite of the youth of {\tt itools} (version {\tt 0.3} at the time of
this writing), it is already being used in real production projects.
Specially, it is the foundation of the {\bf iKaaro} Content Management System.

Some of the advantages provided by its architecture are:

\begin{itemize}
  \item The whole web site data is stored as a tree of human readable
    files, instead of opaque persistent Python objects. It is possible
    to export a web site to the file system, to inspect and manipulate
    the tree, to make a tarball, to import it back to the ZODB.

  \item This approach encourages the use of standards, every time we
    need a new feature we look first if there is already an standard
    file format. The result is a highly portable, standards compliant
    application, easy to interact with other applications.

  \item It is possible to seamlessly integrate content from foreign
    sources and to make it appear as if they were in our web site.
    We are just at the surface of the possibilities.
\end{itemize}

But the most important, it provides a simple, high-productive framework
to build our applications on, to reduce the development time to the
minimum.

\section{Future works}

There is much to do however:

\begin{itemize}
  \item {\tt itools.uri} is not yet 100\% compliant with the RFC2396,
    once mature enough this module even could become part of the
    Python's Standard Library;

  \item support for new schemes, protocols and storages should be
    added to {\tt itools.resources};

  \item there are a thousand handlers to develop, for standard file formats
    like vCard, iCal, etc. 
\end{itemize}

And, last but not least, {\tt itools} is free software, released to the
world under the terms and conditions of the GNU Lesser General Public
License, developers own the copyright of the code they write, contributions
are very welcomed.


\begin{appendix}
\input{arch.tex}
\end{appendix}



\end{document}